### 2021-07-05 rad-icfp-2021/generalisation.tex
Intro until section 1.4 + US spelling
more in gen section

There are two main differences with fold left !foldl! copmpared to !reduce!. 
First, in !foldl (x,y.e) v A! the starting accumulation element !v! is not a unit for !(x,y.e)!,
so it will have non trivial derivatives in general and we need to account for that.
Second, we will need a more general !scanl! which allows !(x,y.e)! as a function argument. 
In other words, we need the general scan left computing the intermediate values of a fold left.
This should be a different primitive but we will still call it !scanl! in this section.

@@

The problem is that this could break the complexity of reverse mode because of the non linear usage of $Y$, and makes everything harder to optimise.

A slighly better option would be to define

@@

It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
but there is still a non linear usage of the continuation variable !Y!.
One solution, if we know we want to compute the whole gradient of the expression, is to define instead

@@

TODO: introduce sum and tensor of semantics of ground type\\
TODO: then * becomes a dot product\\
TODO: minor changes from there\\
TODO: except I'm not sure with reduce and shift1...

Though this has the right complexity, it is an interesting avenue for future research to find
better representations to allow for more optimisations. 
In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.

TODO: more generally about tensors \cite{liao2019differentiable,bernstein2020differentiating}

>>>>

There are two main differences with fold left !foldl! compared to !reduce!. 
First, in !foldl (x,y.e) v A! the starting accumulation element !v! is not a unit for !(x,y.e)!,
so it will have non-trivial derivatives in general and we need to account for that.
Second, we will need a more general !scanl! which allows !(x,y.e)! as a function argument. 
In other words, we need the general scan left computing the intermediate values of a fold left.
This should be a different primitive but we will still call it !scanl! in this section.

@@

The problem is that this could break the complexity of reverse mode because of the non-linear usage of $Y$, and makes everything harder to optimize.

A slightly better option would be to define

@@

It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
but there is still a non-linear usage of the continuation variable !Y!.
One solution, if we know we want to compute the whole gradient of the expression, is to define instead

@@

A ground type $G$ is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context $\Gamma=x1:G1,...,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.

We can define $\otimes$ on the types of our language inductively by

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A1 \times ... \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times ... \times (An \otimes A)$ \\
    $\Array{A1} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times...\times An$ as expected.
TODO: define *, finish D(reduce)

We will need a new primitive $\widehat{*}$ whose denotation is this map.

Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!.

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !map2 (x,y.e1: G) e2 e3!) &=&  
    !let A,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y2 = !$\directD{\rho}{\Gamma,A}{Y1}$!(e3) in! \\
    && !let C=!$\grad_{\Gamma}$!e1 in!\\
    && !<map2 (x,y.e1) A B, fun (x1,...,xn,Z) -> !\\
    && !Y2( (x1,...,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e1)[a/x]!$\widehat{*}$!b) B Z )>!\\
    \end{tabular}
    \end{center}

\begin{center}
\begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
        && !let A0 = scanl (x,y.e1) y1 A in! \\
        && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
        && !let A2 = shift1 (map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A)! \\
        && !let A3 = scanr * 1 A1! \\
        && !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
        && !Y((x1,...,xn),0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

Evidently, one can combine all the generalizations from the previous subsections.
Even though this transformation has the correct complexity, it is open for future research to find
even better representations to allow for more optimizations. 
In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.
